The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-28 paper.tex
Revisao Secao 4.3, 4.4

Each line of execution in \CEU carries associated data, such as which event it 
is awaiting and which code to execute when it awakes.
%
The compiler statically infers the maximum number of trails a program can have 
at the same time and creates a static vector to hold the runtime information 
about them.
%
Trails that cannot be active at the same time can share memory slots in the 
static vector.

@@

All terminated or not-yet-started trails stay in the \code{INACTIVE} state and 
are ignored by the scheduler.
%
A \code{STACKED} trail holds its associated stack level and is delayed until 
the scheduler runtime level reaches that value again.
%
A \code{FIN} trail represents a hanged finalization block which is only 
scheduled when its corresponding block goes out of scope.
%
A trail waiting for an event stays in the state of the corresponding event, 
also holding the minimum sequence number (\emph{seqno}) in which it can awake.
%
A trail is represented by the following \code{struct}:

@@

The size of \code{state\_t} depends on the number of events in the application;
for an application with less than 253 events (plus the 3 states), one byte is 
enough.
%
The size of \code{label\_t} depends primarily on the number of \code{await} 
statements in the application---each \code{await} splits the code in two and 
requires a unique entry point in the code for its continuation.
Additionally, split \& join points for parallel compositions, \code{emit} 
continuations, and finalization blocks also require labels.
%
The \code{seqno} will eventually overflow during execution (every 256 
reactions).
However, given that the scheduler traverses all trails in each reaction, it can 
adjust them to properly handle overflows (actually 2 bits to hold the 
\code{seqno} is already enough).
%
The stack size depends on the maximum depth of nested emissions and is bounded 
to the maximum number of trails, e.g., a trail emits an event that awakes 
another trail, which emits an event that awakes another trail, and so on---the 
last trail cannot awake any trail, because they will be all hanged in a 
\code{STACKED} state.

In the context of embedded systems, the size of \code{trail\_t} is typically 
only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
even for trails that only await a single event and terminate.
%
For instance, the \emph{CTP} collection protocol ported to \CEU has at most 
eight simultaneous lines of execution with an overhead of 2\% in comparison to 
the original version in \emph{nesC} (a dialect of \emph{C} for event-driven 
programming)~\cite{ceu.sensys13}.

@@

In the final generated code in \emph{C}, each trail label representing an entry 
point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails, e.g., the label \code{Awake\_e} will execute on 
\code{TRAIL-0} (line 7).
%
For each statement that yields (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (lines 
1--10, in the middle of the figure).
The trails translate to a vector of \code{trail\_t} elements with the maximum 
number of simultaneous trails (lines 12--16).
Initially, \code{TRAIL-0} is set to execute the \code{Main} entry point, while 
all others are set to \code{INACTIVE}.

The scheduler executes in two passes:
%
In the \emph{broadcast} pass, the scheduler sets to \code{STACKED} all trails 
that are waiting for the current event to execute in the current stack level.
%
In the \emph{dispatch} pass, the scheduler executes each trail that is 
\code{STACKED} to run in the current level, setting it immediately to 
\code{INACTIVE}.

During the dispatch pass, if a trail executes and emits and internal event, the 
scheduler increments the stack level and re-executes the two passes.
After all trails are properly dispatched, the scheduler decrements the stack 
level and resumes the previous execution.
For the first reaction chain, the scheduler starts from the \emph{dispatch} 
pass, given that the \code{Main} label is the only one that can be active at 
the stack level 0 (line 13, in the middle of Figure~\ref{lst.impl.trails}).

The code the right of the Figure~\ref{lst.impl.trails} dispatches a trail
according to its current label to execute.
%
For the first reaction, it executes the \code{Main} label in \code{TRAIL-0}.
%
When the \code{Main} label reaches the \code{par/and}, it ``stacks'' 
\code{TRAIL-1} (lines 4--8) and proceeds to the code in \code{TRAIL-0} (lines 
9--14), respecting the deterministic execution order.
%
The code sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
\code{Awake\_e}, and then halts with a \code{break}.
%
The next iteration of \code{dispatch} takes \code{TRAIL-1} and executes its 
registered label \code{And\_sub\_2} (lines 16--21), which sets \code{TRAIL-1} 
to await \code{EVT\_A} and also halts.

Regarding abortion and finalization, when a \code{par/or} terminates, the 
scheduler makes a \emph{broadcast} pass for the \code{FIN} event, but limited 
to the range of trails covered by the terminating \code{par/or}.
Trails that do not match the \code{FIN} are set to \code{INACTIVE}, as they 
have to be aborted.
Given that trails in parallel are allocated in subsequent slots in the static 
vector \code{TRLS}, this pass only aborts the desirable trails.
The subsequent \emph{dispatch} pass executes the finalization code.
%Given that finalization blocks cannot contain \code{await} statements, the 
%whole process is guaranteed to terminate in bounded time.
Escaping a \code{loop} that contains parallel compositions also trigger the 
same abortion process.

>>>>

Each line of execution in \CEU needs to carry associated data, such as which 
event it is awaiting and which code to execute when it awakes.
%
The compiler statically infers the maximum number of trails a program can have 
at the same time and creates a static vector to hold the runtime information 
about them.
%
Like normal variables, trails that cannot be active at the same time can share 
slots in the static memory vector.

@@

All terminated or not-yet-started trails stay in the \code{INACTIVE} state and 
are ignored by the scheduler.
%
A \code{STACKED} trail holds an associated stack level and is delayed until the 
scheduler runtime reaches that level again.
%
A \code{FIN} trail represents a hanged finalization block which is only 
scheduled when its corresponding block goes out of scope.
%
A trail waiting for an event stays in the state of the corresponding event, 
also holding the minimum sequence number (\emph{seqno}) in which it can awake.
%
In concrete terms, a trail is represented by the following \code{struct}:

@@

The size of \code{state\_t} depends on the number of events in the application;
for an application with less than 253 events (plus the 3 states), one byte is 
enough.
%
The size of \code{label\_t} depends primarily on the number of \code{await} 
statements in the application---each \code{await} splits the code in two 
segments and requires a unique entry point in the code for its continuation.
Additionally, split \& join points for parallel compositions, \code{emit} 
continuations, and finalization blocks also require labels.
%
The \code{seqno} could eventually overflow during execution (i.e., every 256 
reactions).
However, given that the scheduler traverses all trails on every reaction, it 
can adjust them to properly handle overflows (actually, 2 bits to hold the 
\code{seqno} is already enough).
%
The size of \code{stack\_t} depends on the maximum depth of nested emissions 
and is bounded to the maximum number of trails.
In the worst case, a trail emits an event that awakes another trail, which 
emits an event that awakes another trail, and so on.
The last trail cannot awake any trail, because they are all hanged in the 
\code{STACKED} state.

In the context of embedded systems, the size of \code{trail\_t} is typically 
only 3 bytes (1 byte for each field), imposing a negligible memory overhead 
even for trails that only await a single event and terminate.
%
For instance, the \emph{CTP} collection protocol ported to \CEU reaches eight 
simultaneous lines of execution with an overhead of 2\% in comparison to the 
original version in \emph{nesC} (a dialect of \emph{C} for event-driven 
programming)~\cite{ceu.sensys13}.

@@

In the final generated code in \emph{C}, each trail label representing an entry 
point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails, e.g., the label \code{Awake\_e} will execute on 
\code{TRAIL-0} (line 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (lines 
1--10, in the middle of the figure).
The state of trails translate to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (lines 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(line 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in line 14).

The scheduler executes in two passes:
%
In the \emph{broadcast} pass, the scheduler sets all trails that are waiting 
for the current event to \code{STACKED} in the current stack level.
%
In the \emph{dispatch} pass, the scheduler executes each trail that is 
\code{STACKED} to run in the current level, setting it immediately to 
\code{INACTIVE}.

During the dispatch pass, if a trail executes and emits an internal event, the 
scheduler increments the stack level and re-executes the two passes.
After all trails are properly dispatched, the scheduler decrements the stack 
level and resumes the previous execution.
For the first reaction chain, the scheduler starts from the \emph{dispatch} 
pass, given that the \code{Main} label is the only one that can be active at 
the stack level 0 (line 13, in the middle of Figure~\ref{lst.impl.trails}).

The code the right of the Figure~\ref{lst.impl.trails} dispatches a trail
according to the current label to execute.
%
For the first reaction, it executes the \code{Main} label in \code{TRAIL-0}.
%
When the \code{Main} label reaches the \code{par/and}, it stacks \code{TRAIL-1} 
(lines 4--7) and proceeds to the code in \code{TRAIL-0} (lines 9--14), 
respecting the deterministic execution order.
%
The code sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
\code{Awake\_e}, and then halts with a \code{break}.
%
The next iteration of \code{dispatch} takes \code{TRAIL-1} and executes its 
registered label \code{And\_sub\_2} (lines 16--21), which sets \code{TRAIL-1} 
to await \code{EVT\_A} and also halts.

Regarding abortion and finalization, when a \code{par/or} terminates, the 
scheduler makes a \emph{broadcast} pass for the \code{FIN} event, but limited 
to the range of trails covered by the terminating \code{par/or}.
Trails that do not match the \code{FIN} are set to \code{INACTIVE}, as they 
have to be aborted.
Given that trails in parallel are allocated in subsequent slots in the static 
vector \code{TRLS}, this pass only aborts the desirable trails.
The subsequent \emph{dispatch} pass executes the finalization code.
%Given that finalization blocks cannot contain \code{await} statements, the 
%whole process is guaranteed to terminate in bounded time.
Escaping a \code{loop} that contains parallel compositions also triggers the 
same abortion process.
